First paper - Draft

### 2017-07-05 paper/paper.Rnw
Fixed all graphs but table has wrong edu levels. Next gonne deleted and only add avg performance

The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of the children being and schools can do so much to close it. Moreover, these inequalities can be narrowed significantly if the child coming from unfavourable conditions is exposed to high quality learning environments. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\). Although this sounds logical, it's implications are not assesed properly.

@@

It is incomplete to tusy educational inequalities without considering the previlance of intergenerational transmission of human and social capital. Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that education is by far the investment that yields the best returns. This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. We've become so intertwined with the subject that we've virtually spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities. We still don't know, in comparative terms, which countries have big or small cognitive gaps resulting from SES origins. And even more importantly, we haven't assesed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.

@@

A plotting ot the standard deviation of the data suggests that it's something we shouldn't be deeply concerned with. The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.

@@

% Show the properties of the ESCS dummy variable
% % in high educated
% % of home possessions
% % of ISEI

@@

% Figure out how to ouput data frames as latex tables.
<<sample_size>>=
# Get sample counts for each dummy
sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {

@@

We start by look at the achievement for some countries in figure 1. We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden have very steep slopes. France increasde the gap by roughly \Sexpr{diff_data$math[["France"]]} standard deviations, Austria by \Sexpr{diff_data$math[["Austria"]]} and Sweden by \Sexpr{diff_data$math[["Sweden"]]}. This pattern happens similarly for literacy. For example, France has an increase of \Sexpr{diff_data$read[["France"]]}. For such a short period of time, the magnitud of these increases are reasonably big.

@@

Figure 2 below takes a more direct approach and looks at the percentage change from the first and last timepoint available. Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.

@@

Figure 3 shows the evolution for both Low and High SES groups separetly. The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other. We find plenty of variation in these patterns. For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up. The U.K seems to be following the same path as the United States as well. On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster. The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different. We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.

@@

Ths source of this gap can be deconstructed further into two broader gaps. We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group. For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th? Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier? Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap. Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous graphs \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.

% This plot might not be necessary any more
% <<>>=
% 
% lower <- seq(13, 1, -1)
% upper <- seq(100, 220, 10)
% 
% df <- tibble(country = "fake_country",
%              years = rep(seq(2000, 2025, 2), 2),
%              ses = rep(c("lower", "upper"), each = 13),
%              value = c(lower, upper))
% 
% # You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
% df %>%
%   ggplot(aes(as.factor(years), value, group = ses, colour = ses)) +
%   geom_point(size = 0.5, alpha = 0.4) +
%   geom_line(alpha = 0.4) +
%   geom_line(stat = "smooth", method = "lm", aes(group = 1),
%             formula = y ~ splines::ns(x, 3), linetype = "longdash",
%             colour = "black") +
%   # geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
%   # geom_hline(yintercept = 0, linetype = "longdash") +
%   # coord_cartesian(ylim = c(0, 3)) +
%   facet_wrap(~ country)
% @

@@

# Comparing all gaps across countries
complete_data_topbottom %>%
  bind_rows(complete_data_topmid) %>%
  bind_rows(complete_data_midbottom) %>%
  filter(country %in% c("United States", "Denmark", "France", "Germany"),
         type_test == "math") %>%
  mutate(type = factor(type,
                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
                       ordered = TRUE)) %>%
  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test, shape = type_test)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  geom_line(alpha = 0.4) +
  geom_point(size = 0.5, alpha = 0.4) +
  coord_cartesian(ylim = c(0, 3)) +
  scale_y_continuous(name = "Standardized test scores (mean 0)") +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  scale_colour_discrete(name = NULL, labels = c("Low SES", "High SES")) +
  ggtitle("Comparison of the achievement gap for three socio-economic groups") +
  facet_grid(country ~ type) +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey"))
@
\end{center}
\end{figure}

We start off with Denmark. Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger for both tests. Furthermore, we find, just as in the 90th/10th SES gap, that there isn't a significant decrease, with the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation. France shows a similar pattern as Denmark in that the 50th/10th SES gap is the main contributer and that the three gaps have the similar trend.

Germany and U.S, on the other hand, show a very different landscape. For example, Germany, shows that the 90th/10th gap is primarily driven by the 50th/10th SES gap and that the slopes of the 90th/50th gap is much weaker than the other two gaps. This means that Top/Middle SES groups have shrunk much less than the other groups. The big catching up is actually coming from the Middle/Bottom SES groups. However, we don't know if this is because the bottom is catching up or that the middle is coming down. However, based on the gap plots from above, we can infer that it is because the bottom is catching up rapidly because against the 90th SES group, it is increasing steeply every year. The U.S shows a different pattern from all other countries. Both the 90th/50th/10th gaps seem to contribute rather similarly to the overall 90th/10th gap (although 50th/10th is marginally higher) but more interestingly, the 90th/50th gap has stagnated in the last 15 years. The top/middle groups have a relatively big gap and it hasn't changed at all. On the other hand, the steep decrease we saw for the U.S is coming exclusively from the 50th/10th gap shrinking. The reason, nonetheless, might be different from the German case, where it's probable that the bottom is catching up. Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group.

In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing.

@@

full_data %>%
  ggplot(aes(low_increase, high_increase), alpha = 0.2) +
  geom_rect(data = rect_data, aes(xmin = xst,
                                  xmax = xen,
                                  ymin = yst,
                                  ymax = yen),
            fill = rect_data$colour,
            alpha = 0.2,
            inherit.aes = FALSE) +
  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1,
            linetype = "longdash") +
  geom_point(alpha = 0.2) +
  geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
  geom_line(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.4,
            stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "red", size = 1) +
  geom_text_repel(data = filter(full_data, continent == "my_cnt"),
                  aes(label = country), box.padding = unit(2.7, "lines")) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  xlim(lims$xlim) +
  ylim(lims$ylim) +
  coord_cartesian(expand = FALSE) +
  annotate(geom = "text", x = 0.15, y = -0.2,
           label = "Low SES is catching up \n faster than High SES",
           fontface = 2, size = 3) +
  annotate(geom = "text", x = -0.05, y = 0.20,
           label = "High SES is increasing  \n faster than Low SES is catching up",
           fontface = 2, size = 3) +
  labs(x = "Average increase of low SES in SD", y = "Average increase of high SES in SD") +
  theme_minimal()
@
\end{center}
\end{figure}

We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as United States and Germany. We see a negative relationship where the faster the low SES catch up the slower the High SES decrease. Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more ocuntries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps below.

@@

Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's an efficiency/equality tradeoff and if it stands over time. One particular topic that has been researched in the past is whether improving the overall cognitive development of a country compromises the distribution of the countries overall performance. For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance. This, however, can come at a price. If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.

@@

\begin{figure}
\begin{center}
<<avgperformance_inequality, include = T, out.height = '5in', out.width = '7in', fig.align = 'center'>>=
data_to_plot %>%
  ggplot(aes(scale(Mean), difference)) +
  geom_smooth(method = "lm") +
  geom_point() +
  scale_y_continuous(name = "90/10 achievement gap") +
  scale_x_continuous(name = "Average performance standardized") +
  ggtitle("Relationship between average performance and achievement gap") +
  facet_wrap(~ wave, nrow = 2) +
  theme_few()
@
\end{center}
\end{figure}

For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}. The correlations are between \Sexpr{max(correlation)} and \Sexpr{min(correlation)}. This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}. However, what they didn't really go in depth was whether this corelation is strongest for some countries over time. With the data available we can look at the evolution separately for many countries.

\begin{figure}
\begin{center}
<<country_achievement_and_inequality, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
data_to_plot %>%
  mutate(Mean = scale(Mean),
         country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  gather(type, value, difference, Mean) %>%
  filter(country %in% countries) %>%
  ggplot(aes(as.character(wave), value, group = type, colour = type)) +
  geom_point(alpha = 0.4) +
  geom_line(alpha = 0.4) +
  scale_y_continuous(name = "Standardized test scores (mean 0)") +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  scale_colour_discrete(name = NULL, labels = c("Achievement gap", "Avg performance")) +
  ggtitle("Evolution of the top/bottom groups by different gaps") +
  facet_wrap(~ country, ncol = 5) +
  theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        legend.position = "bottom")
@
\end{center}
\end{figure}

In fact, we see a strong correspondence. In countries like France we see quite a functional form between both inequality and performance. As average performance plummets, the achievement gap is increasing. This pattern is also seen in in the Austria, Australia and Germany. The other obvious pattern is the opposite trend. We see countries like Poland, Finland, Canada and much less pronounced Italy that this pattern is present. The surprising case here is the United States. We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well. In fact, among the few countries we should find something that resembles this pattern should be the U.S because it doesn't have a strict institutionalized tracking system. However, the explanation behind this pattern is very speculative. Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap. Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality.

>>>>

The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of a child begins and schools can do so much to close it. These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\). Although this sounds logical, it's importance is not understood entirely.

@@

It is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that education is by far the investment that yields the best returns. This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities. We still don't know, in comparative terms, which countries have big or small cognitive gaps resulting from SES origins. And even more importantly, we haven't assesed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.

@@

We plot the standard deviation of the mathematics test score for all waves in figure 1. The plot suggests that it's something we shouldn't be deeply concerned with. The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.

@@

@@

<<sample_size>>=
# Get sample counts for each dummy
sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {

@@

We start by look at the achievement for some countries in figure 2. We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden have very steep slopes. France increasde the gap by roughly \Sexpr{diff_data$math[["France"]]} standard deviations, Austria by \Sexpr{diff_data$math[["Austria"]]} and Sweden by \Sexpr{diff_data$math[["Sweden"]]}. This pattern happens similarly for literacy. For example, France has an increase of \Sexpr{diff_data$read[["France"]]}. For such a short period of time, the magnitud of these increases are reasonably big.

@@

Figure 3 below takes a more direct approach and looks at the percentage change from the first and last timepoint available. Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.

@@

Figure 4 shows the evolution for both Low and High SES groups separetly. The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other. We find plenty of variation in these patterns. For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up. The U.K seems to be following the same path as the United States as well. On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster. The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different. We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.

@@

Ths source of this gap can be deconstructed further into two broader gaps. We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group. For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th? Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier? Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap in figure 5. Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous graphs \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.

@@

# Comparing all gaps across countries
complete_data_topbottom %>%
  bind_rows(complete_data_topmid) %>%
  bind_rows(complete_data_midbottom) %>%
  filter(country %in% c("United States", "Denmark", "France", "Germany"),
         type_test == "math") %>%
  mutate(type = factor(type,
                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
                       ordered = TRUE)) %>%
  ggplot(aes(as.factor(wave), difference, group = type_test)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  geom_linerange(aes(ymin = 0, ymax = difference)) +
  geom_line(stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 2), size = 0.7,
            colour = "red") +
  geom_point(size = 0.5, alpha = 0.4) +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  coord_cartesian(ylim = c(0, 3)) +
  ggtitle("Comparison of the achievement gap for three socio-economic groups") +
  facet_grid(country ~ type) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"))
@
\caption{Evolution of 90/50/10 SES gaps for selected countries}
\end{center}
\end{figure}

We start off with Denmark. Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger in mathematics. Furthermore, we find, just as in the 90th/10th SES gap, that there isn't a significant decrease \footnote{With the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation.}. France shows a similar pattern as Denmark in that the 50th/10th SES gap is the main contributer and that the three gaps have the similar trend.

Germany and U.S, on the other hand, show a very different landscape. For example, Germany, shows that the 90th/10th gap is primarily driven by the 50th/10th SES gap and that the slopes of the 90th/50th gap is much weaker than the other two gaps. This means that top/middle SES groups have shrunk much less than the other groups. The big catching up is actually coming from the middle/bottom SES groups. However, we don't know if this is because the bottom is catching up or that the middle is coming down. Based on the gap plots from above, we can infer that it is because the bottom is catching up rapidly because against the 90th SES group, it is increasing steeply every year.

The U.S shows a different pattern from all other countries. Both the 90th/50th/10th gaps seem to contribute rather similarly to the overall 90th/10th gap (although 50th/10th is marginally higher) but more interestingly, the 90th/50th gap has stagnated in the last 15 years. The top/middle groups have a relatively big gap and it hasn't changed at all. On the other hand, the steep decrease we saw for the U.S is coming exclusively from the 50th/10th gap shrinking. The reason, nonetheless, might be different from the German case, where it's probable that the bottom is catching up. Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group. The big takeaway from the U.S is that the decrease of inequality is coming from the shrinking of the 50/10 gap.

In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing in figure 6.

@@

full_data %>%
  ggplot(aes(low_increase, high_increase), alpha = 0.2) +
  geom_rect(data = rect_data, aes(xmin = xst,
                                  xmax = xen,
                                  ymin = yst,
                                  ymax = yen),
            fill = rect_data$colour,
            alpha = 0.2,
            inherit.aes = FALSE) +
  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1,
            linetype = "longdash") +
  geom_point(alpha = 0.2) +
  geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
  geom_line(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.4,
            stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "red", size = 1) +
  geom_text_repel(data = filter(full_data, continent == "my_cnt"),
                  aes(label = country), box.padding = unit(2.7, "lines")) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  xlim(lims$xlim) +
  ylim(lims$ylim) +
  coord_cartesian(expand = FALSE) +
  annotate(geom = "text", x = 0.15, y = -0.2,
           label = "Low SES is catching up \n faster than High SES",
           fontface = 2, size = 3) +
  annotate(geom = "text", x = -0.05, y = 0.20,
           label = "High SES is increasing  \n faster than Low SES is catching up",
           fontface = 2, size = 3) +
  labs(x = "Average increase of low SES in SD", y = "Average increase of high SES in SD") +
  theme_minimal()
@
\caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
\end{center}
\end{figure}

We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany. We see a negative relationship where the faster the low SES catch up the slower the High SES decrease. This is very interesting because it suggests that there's a trend for countries to have the bottom SES groups are climbing upwards, a trait of meritocracy. Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more ocuntries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps in figure 7.

@@

Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's an efficiency/equality tradeoff and if it stands over time. One particular topic that has been researched in the past is whether improving the overall cognitive development of a country compromises the distribution of the countries overall performance. For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance. This, however, can come at a price. If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance. We plot this relationship in figure 8.

@@

\begin{figure}
\begin{center}
<<avgperformance_inequality, include = T, out.height = '5in', out.width = '6.5in', fig.align = 'center'>>=
data_to_plot %>%
  ggplot(aes(scale(Mean), difference)) +
  geom_smooth(method = "lm") +
  geom_point() +
  scale_y_continuous(name = "90/10 achievement gap") +
  scale_x_continuous(name = "Average performance standardized") +
  ggtitle("Relationship between average performance and achievement gap") +
  facet_wrap(~ wave, nrow = 2) +
  theme_few()
@
\caption{Relationship between 90/10 achievement gap and average performance in mathematics}
\end{center}
\end{figure}

For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}. The correlations are between \Sexpr{max(correlation)} and \Sexpr{min(correlation)}. This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}. However, what they didn't really go in depth was whether this corelation is strongest for some countries over time. With the data available we can look at the evolution separately for many countries. We plot it up next in figure 9.

\begin{figure}
\begin{center}
<<country_achievement_and_inequality, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
data_to_plot %>%
  mutate(Mean = scale(Mean),
         country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  gather(type, value, difference, Mean) %>%
  filter(country %in% countries) %>%
  ggplot(aes(as.character(wave), value, group = type, colour = type, shape = type)) +
  geom_point(alpha = 0.4) +
  geom_line(alpha = 0.4) +
  scale_y_continuous(name = "Standardized test scores (mean 0)") +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  scale_colour_discrete(name = NULL, labels = c("Achievement gap", "Avg performance")) +
  scale_shape_discrete(name = NULL, labels = c("Achievement gap", "Avg performance")) +
  ggtitle("Evolution of the top/bottom groups by different gaps") +
  facet_wrap(~ country, ncol = 5) +
  theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        legend.position = "bottom")
@
\caption{Time trends of achievement gap and average achiement between 2000 and 2015}
\end{center}
\end{figure}

Looking at figure 9, we see a strong correspondence between the two variables. In countries like France we see quite a functional form between both inequality and performance. As average performance plummets, the achievement gap is increasing. This pattern is also seen in Austria, Australia and Germany. The other obvious pattern is the opposite trend. We see countries like Poland, Finland, Canada and less pronounced, Italy, where this pattern is evident. The surprising case here is the United States. We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well. In fact, among the few countries we should find something that resembles this pattern should be the U.S because it doesn't have a strict institutionalized tracking system. However, the explanation behind this pattern is very speculative.

Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap. Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentially, are countries that have the lower leven of inequalities among the ones presented here.
